Feasibility of Harvesting Power to Run a Domestic Water Meter Using Streaming Cell Technology

### 2014-11-24 paper.tex
Final submission

\title{Feasibility of Harvesting Power To Run A Domestic Water Meter Using Streaming Cell Technology}

@@

    The remainder of this work is presented as follows.
    In section~\ref{sect:streamingCell} we build, measure, and calculate the conversion efficiency of a simple streaming potential cell.
    We confirm the relationship between voltage and applied pressure and investigate the effect of channel height.
    In section~\ref{sect:waterConsumption} we estimate the amount of harvestable energy from a domestic water usage profile.
    Section~\ref{sect:powerRequirements} quantifies energy requirements of a simple electronic meter reader with a wireless transmitter.
    In section~\ref{sect:harvesterSize} we calculate the physical size of a harvester suitable for automatic meter reading.
    And finally, we conclude with a discussion on the feasibility of using streaming cells to harvest energy.

@@

    In~\cite{Gu2000}, the authors employ a relatively simple parallel plate design to create a streaming cell.
    They glued plastic shims between parallel plates of glass, providing a simple way of setting a channel's height.
    Using that method, the authors fabricated three cells with internal channel heights of \SI{50}{\micro\metre}, \SI{100}{\micro\metre} and \SI{150}{\micro\metre}.
    Each of the channels were \SI{3}{\centi\metre} long and had a width of \SI{1}{\centi\metre}.

@@

    Based on the pressure applied and the flow rate through the cell, the input power is \SI{539}{\milli\watt}.
    Energy conversion efficiency is in the order of \SI{1}{\micro\percent}.
    While the conversion efficiency is much lower than those stated in the introduction, we use much larger channel dimensions and tap water as our working fluid.

@@

    \section{Sizing of Streaming Cell Harvester}
    \label{sect:harvesterSize}
    We now estimate the physical size of a harvester capable of supplying the energy to run a wireless water meter.
    Auckland tap water has a sodium content of 2.1--\SI{26.0}{\milli\gram\per\litre}.~\cite{WatercareNewZealand2012}
    This equates to a Debye length in the range 0.3--\SI{1.0}{\nano\metre} and therefore the optimum channel height between 0.6--\SI{2.1}{\nano\metre}.
    We envision a harvester made from stacked layers of glass, each layer forming a rectangular channel similar to those constructed earlier.

    Such a harvester would be installed as shown in Fig.~\ref{fig:harvester} to control the pressure loss within the harvester.
    Using the head loss curve from Fig.~\ref{fig:headloss}, a flow rate of \SI{7.50}{\litre\per\minute} would produce a pressure differential of \SI{4.5}{\kilo\pascal}.
    Fig.~\ref{fig:cellVoltagePressure} confirms that the harvester output is linearly proportional to the pressure applied.

@@

    We know from Section~\ref{sect:waterConsumption} that a typical shower consumes \SI{7.50}{\litre\per\minute} for approximately \SI{396}{\second}, dissipating a total of \SI{222}{\joule} of energy inside a water meter.
    This equates to a power dissipation of \SI{560}{\milli\watt}.

    This means our harvester must be capable of harvesting at least \SI{5.6}{\milli\watt} for a water flow rate of \SI{7.50}{\litre\per\minute}.

    From the dimensions of the cell constructed and measured in Section~\ref{sect:streamingCell} we can expect \SI{152}{\nano\watt\per\metre} of channel width.
    We assume here that doubling the width of a channel will double the output power of the channel.

    To increase the efficiency of the channel the internal height needs to be reduced by a factor of \SI{52600}, from \SI{71}{\micro\metre} to \SI{1.35}{\nano\metre} (the midpoint between \SI{0.6}{\nano\metre} and \SI{2.1}{\nano\metre}).
    Doing so decreases the normalised output power of the streaming cell to \SI{2.89}{\pico\watt\per\metre}.
    Therefore the width of a channel required to generate the target \SI{5.6}{\milli\watt} is \SI{1.94}{\giga\metre}.

    Building such a device could be achieved by stacking \SI{1970000} glass channels, each of width \SI{984}{\metre} and thickness \SI{500}{\micro\metre}, creating a harvester that is \SI{968}{\square\kilo\metre} in size.
    Keeping the channel length at \SI{1}{\centi\meter}, it may be possible to fit the required channels into a \SI{9.68}{\cubic\kilo\meter} volume.

@@

    \section{Conclusion}
    \label{sect:conclusion}
    Approximately one hundred times the required power to operate an automatic meter reader and transmitter is lost in current mechanical water meters.
    The streaming cells we constructed produced conversion efficiencies in the order of \SI{1}{\micro\percent}.
    Scaling these cells to create a viable harvester would result in a \SI{9.68}{\cubic\kilo\metre} volume of glass with a mass of \SI{24.4}{\giga\gram}.
    This is far from being a practical solution to the battery operated devices currently in use.

    Theoretical methods proposed in the literature to increase the efficiency of streaming cells would reduce these figures.
    Regardless of efficiency, it is expected that such narrow openings in domestic water feeds will eventually become clogged.

    It is our conclusion that energy harvesting using streaming cells for the purpose of automatic meter reading is not practical.
    Based on the quantity of energy already dissipated within mechanical meters, other methods of energy harvesting should produce favourable results.
    \bibliographystyle{ieeetr}
    \bibliography{library}

>>>>

\title{Feasibility of Harvesting Power to Run a Domestic Water Meter Using Streaming Cell Technology}

@@

        The remainder of this work is presented as follows.
        In section~\ref{sect:streamingCell} we build, measure, and calculate the conversion efficiency of a simple streaming potential cell.
        We confirm the relationship between voltage and applied pressure and investigate the effect of channel height.
        In section~\ref{sect:waterConsumption} we estimate the amount of harvestable energy from a domestic water usage profile.
        Section~\ref{sect:powerRequirements} quantifies energy requirements of a simple electronic meter reader with a wireless transmitter.
        In section~\ref{sect:harvesterSize} we calculate the physical size of a harvester intended for electronic meter reading.
        And finally, we conclude with a discussion on the feasibility of using streaming cells to harvest energy.

@@

        In~\cite{Gu2000}, the authors employ a relatively simple parallel plate design to create a streaming cell.
        They glued plastic shims between parallel plates of glass, providing a simple way of setting a channel's height.
        Using that method, the authors fabricated three cells with internal channel heights of \SI{50}{\micro\metre}, \SI{100}{\micro\metre} and \SI{150}{\micro\metre}.
        Each of the channels were \SI{3}{\centi\metre} long and had an internal width of \SI{1}{\centi\metre}.

@@

        By calculating the power provided to the harvester and knowing the power output we calculate the cell efficiency.

        % \begin{equation}
        %     P_{in} = \SI{260}{kPa} \cdot 2.05\times10^{-6} \frac{m^{3}}{s} = \SI{533}{mW}
        % \end{equation}
        % \begin{equation}
        %     P_{out} = \SI{1.53}{\nano\watt}
        % \end{equation}
        % \begin{equation}
        %     \epsilon = \frac{P_{out}}{P_{in}} = 2.87\times 10^{-9} = 0.287\thinspace\mu\%
        % \end{equation}

        \begin{eqnarray}
            P_{in} =& \SI{260}{kPa} \cdot \SI{2.05e-6}{\cubic\meter\per\second} = \SI{533}{mW}\\
            P_{out} =& \SI{1.53}{\nano\watt}\\
            \epsilon =& \frac{P_{out}}{P_{in}} = 2.87\times 10^{-9} = 0.287\thinspace\mu\%
        \end{eqnarray}

        Energy conversion efficiency is in the order of \SI{0.3}{\micro\percent}.
        While the conversion efficiency is much lower than those stated in the introduction, we use much larger channel dimensions.

@@

    \section{Sizing of Streaming Cell Harvester}
        \label{sect:harvesterSize}
        We now estimate the physical size of a harvester capable of supplying enough energy to run an electronic water meter.
        We envision a harvester made from stacked layers of glass, each layer forming a rectangular channel similar to those constructed earlier.
        Such a harvester would be installed as shown in Fig.~\ref{fig:harvester} to control the pressure loss within the harvester and to the consumer.

@@

        Auckland tap water has a sodium content of 2.1--\SI{26.0}{\milli\gram\per\litre}.~\cite{WatercareNewZealand2012}
        This equates to a Debye length in the range 0.3--\SI{1.0}{\nano\metre} and therefore the optimum channel height between 0.6--\SI{2.1}{\nano\metre}.

        We know from Section~\ref{sect:waterConsumption} that a typical shower consumes \SI{7.50}{\litre\per\minute} for approximately \SI{396}{\second}.
        Using the head loss curve from Fig.~\ref{fig:headloss}, a flow rate of \SI{7.50}{\litre\per\minute} would produce a pressure differential of \SI{4.48}{\kilo\pascal}.
        Therefore, a mechanical water meter dissipates a total of \SI{222}{\joule} at a rate of \SI{560}{\milli\watt} within the meter during a typical shower.

        This means a harvester must be capable of harvesting at least \SI{5.60}{\milli\watt} at a pressure differential of \SI{4.48}{\kilo\pascal}.
        From the dimensions of the cell constructed and measured in Section~\ref{sect:streamingCell} we can expect \SI{153}{\nano\watt\per\metre} of channel width at a pressure of \SI{260}{\kilo\pascal}.
        Fig.~\ref{fig:cellVoltagePressure} confirms that the harvester output is linearly proportional to the pressure applied.
        We assume here that doubling the width of a channel will double the output power of the channel.

        Normalising the cell's output for width and pressure yields \SI{588}{\pico\watt\per\meter\per\kilo\pascal}.
        Therefore at our target pressure of \SI{4.48}{\kilo\pascal} we can expect output of \SI{2.64}{\nano\watt\per\meter}.
        Dividing the target output of \SI{5.60}{\milli\watt} by this figure yields a required width of \SI{2.12e6}{\meter}.
        This equates to a harvester that, when packed into a cube, would have sides of \SI{3.17}{\meter} in length.

        % Setting the glass substrate at thickness \SI{500}{\micro\meter}, and using a channel depth of \SI{3.00}{\centi\meter}, this length can reduced by stacking sections of channels.
        % Stacking these cells on top of and behind each other would then result in a cube having sides of \SI{3.17}{\meter}.
        % This cube would be capable of powering the electronic water meter with a pressure differential of \SI{4.48}{\kilo\pascal}.
        % That pressure differential would be unobtainable using a domestic feed across a channel with those dimensions.

        % Now we consider a harvester with internal channel height of \SI{1.35}{\nano\metre}.
        % We have chosen this dimension as it is the midpoint between \SI{0.6}{\nano\metre} and \SI{2.1}{\nano\metre}, the optimum channel height for the conductivity of tap water.
        % At this dimension we would hope for a conversion efficiency of 1\%, half the theoretically predicted 2\%.

        % Increasing the channel's efficiency to 1\% puts its output at \SI{2.05}{\milli\watt\per\meter\per\kilo\pascal}.
        % However, reducing the channel height from \SI{75}{\micro\meter} to \SI{1.00}{\nano\meter} while preserving channel cross-section drops the power-per-width down to \SI{27.3}{\nano\watt\per\meter\per\kilo\pascal}.
        % At the target pressure of \SI{4.48}{\kilo\pascal} this delivers \SI{122}{\nano\watt\per\meter}.
        % To get \SI{5.60}{\milli\watt} would require a \SI{45.7}{\kilo\meter} wide channel.
        % Again stacking and packing the channels into a cube, the harvester would occupy a \SI{0.229}{\cubic\meter} volume.
        % This is equivalent to a cube having sides of \SI{62}{\centi\meter}.

        % To increase the efficiency of the channel the internal height needs to be reduced by a factor of \SI{52600}, from \SI{71}{\micro\metre} to \SI{1.35}{\nano\metre} (the midpoint between \SI{0.6}{\nano\metre} and \SI{2.1}{\nano\metre}).
        % Doing so decreases the normalised output power of the streaming cell to \SI{2.89}{\pico\watt\per\metre}.
        % Therefore the width of a channel required to generate the target \SI{5.6}{\milli\watt} is \SI{1.94}{\giga\metre}.

        % Building such a device could be achieved by stacking \SI{1970000} glass channels, each of width \SI{984}{\metre} and thickness \SI{500}{\micro\metre}, creating a harvester that is \SI{0.968}{\square\kilo\metre} in size.
        % Keeping the channel length at \SI{1}{\centi\meter}, it may be possible to fit the required channels into a \SI{9680}{\cubic\meter} volume.

@@

    \section{Conclusion}
        \label{sect:conclusion}
        The streaming cells fabricated in this paper achieved a conversion efficiency of approximately \SI{0.3}{\micro\%}.
        Stacking these cells into harvester capable of powering an electronic water meter would result in a cube having sides \SI{3.17}{\meter} in length.

        %Methods proposed to further increase the efficiency of streaming cell energy conversion would reduce this figure.
        We have estimated that harvesting 1\% of the power dissipated in a mechanical water meter would be sufficient to power an electronic water meter.
        In order to achieve this, such a cell would have to operate near the maximum predicted efficiency.
        % A streaming cell harvester capable of achieving the target 1\% conversion efficiency would have to operate near the theoretical maximum predicted in the literature.
        Regardless of efficiency, it is expected that such narrow openings in domestic water feeds will eventually become clogged.
        We conclude that energy harvesting using readily fabricated streaming cells for the purpose of electronic meter reading is not yet practical.

        Methods of inducing hydrodynamic slip at the interface may produce a more feasible streaming cell harvester design.
        %We have estimated that harvesting 1\% of the power already lost inside a mechanical water meter would be sufficient to operate an electronic meter.

        % Based on this figure, we suggest that other methods of energy harvesting should produce more viable results.
    \bibliographystyle{ieeetr}
    \bibliography{library}
